{RandomRationalPoints} package for \emph{Macaulay2}

### 2021-01-22 FastLinAlg/Paper/RandomRationalPoints.tex
Works Now
minor changes

\begin{document}
\title{{RandomRationalPoints} package for \emph{Macaulay2}}
\author{Sankhaneel Bisui}
\email{sbisui@tulane.edu}
\address{Department of Mathematics, Tulane University, New Orleans, LA 70118}
\author{Zhan Jiang}
\email{zoeng@umich.edu}
\address{Department of Mathematics, University of Michigan, Ann Arbor, MI 48109}
\author{Sarasij Maitra}
\email{sm3vg@virginia.edu}
\address{Department of Mathematics, University of Virginia, Charlottesville, VA 22904}
\author{Thai Nguyen}
\email{tnguyen11@tulane.edu}
\address{Department of Mathematics, Tulane University, New Orleans, LA 70118}
\author{Karl Schwede}
\email{schwede@math.utah.edu}
\address{Department of Mathematics, University of Utah, 155 S 1400 E Room 233, Salt Lake City, UT, 84112}
%\date{\today}

@@

    \item {\tt GenericProjection} projects to a hypersurface, via 
        {\tt projectionToHypersurface} and then uses a {\tt BruteForce} strategy.

@@

    \begin{example}\label{BruteForce}
        Consider the following example.
        ~~        
        {\small\color{blue}
    \begin{quote}
\begin{verbatim}
i2 : R = ZZ/101[x, y, z];

i3 : J = ideal(x^3 + y^2 + 1, z^3 - x^2 - y^2 + 2);

i4 : randomPoints(J)
    -- used 0.0181549 seconds

o4 = {{32, 37, -30}}

o4 : List

i5 : time randomPoints(J,Strategy=>BruteForce)
    -- used 0.0146932 seconds

o5 = {}

o5 : List

i6 : randomPoints(J,Strategy=>GenericProjection)
    -- used 0.0146932 seconds

o6 = {{2, -30, -8}}
        \end{verbatim}
    \end{quote}
        }
    \end{example}%
    %
    \vspace{-1em}
    \item[\tt ProjectionAttempts => ZZ]

    When calling the Strategy {\tt GenericProjection} or \\
    {\tt HybridProjectionIntersection} from {\tt randomPoints}, this option denotes the number of trials before giving up.  This option is also passed to {\tt randomPoints} by other functions.

@@

    \begin{example}
        We re-compute \cref{BruteForce} this time specifying more attempts.
    {{\small\color{blue}
    ~~
    \begin{quote}
\begin{verbatim}    
i7 : randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 20000)
    -- used 0.84679 seconds

o7 = {{-39, -43, 28}}
    \end{verbatim}%        
\end{quote}\vspace{-1em}%
}}%
    \end{example}%    
%
    %\vspace{1em}
    \item[\tt MaxCoordinatesToTrivialize]

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.  A quick comment, while in many examples generic projection works well, there are some examples where it performs quite slowly.  Setting {\tt ProjectionAttempts=>0} in the {\tt Default} strategy is a way to avoid this situation.

    \begin{example}
        We begin with an example over a small field.
    {{\small\color{blue}
    ~~
    \begin{quote}
\begin{verbatim}
i2 : R = ZZ/7[x_1..x_10];

i3 : I = ideal(random(2, R), random(3, R));

i4 : time randomPoints(I, Strategy => BruteForce, PointCheckAttempts => 20000)
    -- used 0.0102075 seconds

o4 = {{-2, -1, 3, -2, -2, 2, 0, -2, -2, -2}}

o4 : List

i5 : time randomPoints(I, Strategy => Default)
    -- used 0.085741 seconds

o5 = {{1, -2, 2, 2, -2, 0, 0, -1, -1, 3}}
    \end{verbatim}
\end{quote}\vspace{-1em}
    }}
    \end{example}

\begin{example}
    Now we work over a larger field.
    {{\small\color{blue}
    \begin{quote}
\begin{verbatim}
i6 : S = ZZ/211[x_1..x_10];

i7 : J = ideal(random(2, S), random(3, S));

o7 : Ideal of S

i8 : time randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 2000000)
    -- used 25.3479 seconds

o8 = {{-36, 26, -63, -7, -48, 78, -44, -22, 105, 39}}

i9 : time randomPoints(J, Strategy => Default)
-- used 0.102918 seconds

o9 = {{0, -42, -53, -33, -65, -27, -105, -37, 99, 64}}

o9 : List

i10 : time randomPoints(J, Strategy => LinearIntersection)
-- used 0.0169781 seconds

o10 = {{-51, -11, -34, -33, -65, 19, -86, 50, -44, -97}}

o10 : List

i11 : time randomPoints(J, Strategy => GenericProjection)
-- used 0.364902 seconds

o11 = {{31, -15, 53, 41, 87, 9, -21, -43, -12, 85}}      
    \end{verbatim}
\end{quote}\vspace{-1em}
        }}
    \end{example}

    \begin{example}
        Finally, we allow our functions to extend our field.
        {{\small\color{blue}
        \begin{quote}
        \begin{verbatim}
i12 : time randomPoints(J, Strategy => LinearIntersection, ExtendField => true)
-- used 0.0612306 seconds

            5      4      3      2                                      5  
o12 = {{54, 14a  - 48a  - 56a  + 56a  - 13a + 48, -101, 56, -42, -27, - 26a  -
    -------------------------------------------------------------------------
    4      3      2
    43a  - 70a  - 96a  + 50a - 103, -29, 0, -37}}
        \end{verbatim}
    \end{quote}
        }}
    \end{example}

@@

This function takes a polynomial ring as an input and outputs the coordinate change map,
i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

@@

\begin{itemize}
\item {\tt Replacement}: Setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
\item {\tt MaxCoordinatesToReplace}: The user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
  This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).		
\end{itemize} 
%\par
%\vspace{-1em}    
\begin{example}  We demonstrate some of these options.
%
  {{\small\color{blue}
  \begin{quote}
\begin{verbatim}
	i3 : R = ZZ/11[x, y, z];

	i4 : randomCoordinateChange(R)

           ZZ
o4 = map(R,--[x, y, z],{4x + 5y - 5z, 3x - 4y - 3z, 4x})                     
           11

                    ZZ
o4 : RingMap R <--- --[x, y, z]
                    11

	i5 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1)

	o5 = | x -x-4y-5z y |

	i6 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Homogeneous => false)

	o6 = | x-3 z-5 -x+3y-4z+2 |

	i7 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Replacement => Binomial)

	o7 = | y x+4z z |
\end{verbatim}
\end{quote}
    }}

@@

\begin{example}	
    We project a curve in 4-space to one 2-space.
    ~~
  {{\small\color{blue}
  \begin{quote}
\begin{verbatim}
i1 : R = ZZ/5[x, y, z, w];

i2 : I = ideal(x, y^2, w^3 + x^2);

i3 : genericProjection(2, I)

            ZZ                                         2          2
o3 = (map(R,--[z, w],{- x - 2y - z, - y - 2z}), ideal(z  - z*w - w ))                   
            5
\end{verbatim}
\end{quote}
    }}
\end{example}

Alternatively, instead of {$I$}, we may pass it a quotient ring.  It will then return the inclusion of the generic projection ring into the given ring, followed by the source of that inclusion.  It is essentially the same functionality as calling {\tt genericProjection(n, ideal R)} although the format of the output is slightly different.

@@

i1 : R = ZZ/5[x, y, z, w];

i2 : I = ideal(x, y^2, w^3 + x^2);

i3 : genericProjection(2, R/I)

                            ZZ                          ZZ
                            --[z, w]                    --[z, w]                        
                R           5                           5
o3 = (map(----------------,--------,{2y - w, - y - z}), --------)                       
             2   3    2        4                            4
        (x, y , w  + x )      z                            z

o3 : Sequence
\end{verbatim}}
    }}
\end{comment}

@@

i2 : R = ZZ/5[x, y, z, w];

i3 : I = ideal(x, y^2, w^3 + x^2);

i4 :genericProjection(2, I, Replacement => Binomial)

            ZZ                                   4
o4 = (map(R,--[z, w],{x + 2z, - 2y + w}), ideal w )                                     
            5

o4 : Sequence
\end{verbatim}}
    }}
\end{comment}

@@

    o4 : Sequence

	i5 : projectionToHypersurface(R/I)

@@

i3 : I = ideal(random(3, R) - 2, random(2, R))

                3    2       2     3     2              2       2     3
o3 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z )

i4 : M = jacobian(I)

o4 = {1} | -x2+2xy-y2+xz+2yz    x-2y+z   |
     {1} | x2-2xy-y2+2xz+2yz+z2 -2x-y-2z |
     {1} | -2x2+2xy+y2+2yz-z2   x-2y-2z  |

             3       2
o4 : Matrix R  <--- R

i5 : extendIdealByNonZeroMinor(2, M, I, Strategy => LinearIntersection)

                3    2       2     3     2              2       2     3
o5 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2   3    2        2    3  
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z , x  + x y + 2x*y  - y  +
-------------------------------------------------------------------
  2       2    3
2x z + y*z  - z )

o5 : Ideal of R
\end{verbatim}}	
    }}	
\end{comment}

@@

\bibliographystyle{skalpha}
\bibliography{MainBib}

\end{document}

>>>>

\begin{document}
\title{{RandomRationalPoints} package for \emph{Macaulay2}}
\author{Sankhaneel Bisui}
\email{sbisui@tulane.edu}
\address{Department of Mathematics, Tulane University, New Orleans, LA 70118}
\author{Zhan Jiang}
\email{zoeng@umich.edu}
\address{Department of Mathematics, University of Michigan, Ann Arbor, MI 48109}
\author{Sarasij Maitra}
\email{sm3vg@virginia.edu}
\address{Department of Mathematics, University of Virginia, Charlottesville, VA 22904}
\author{Th\'ai Th\`anh Nguy$\tilde{\text{\^e}}$n}
\email{tnguyen11@tulane.edu}
\address{Department of Mathematics, Tulane University, New Orleans, LA 70118}
\author{Karl Schwede}
\email{schwede@math.utah.edu}
\address{Department of Mathematics, University of Utah, 155 S 1400 E Room 233, Salt Lake City, UT, 84112}
%\date{\today}

@@

    \item {\tt GenericProjection} projects to a hypersurface, via 
        {\tt projectionToHypersurface} and then uses the {\tt BruteForce} strategy.

@@

    \begin{example}\label{BruteForce}
        Consider the following example.
        ~~        
        {\small\color{blue}
    \begin{quote}
\begin{verbatim}
i2 : R = ZZ/101[x, y, z];
i3 : J = ideal(x^3 + y^2 + 1, z^3 - x^2 - y^2 + 2);
o3 : Ideal of R
i4 : randomPoints(J)
    -- used 0.0181549 seconds
o4 = {{32, 37, -30}}
o4 : List
i5 : time randomPoints(J,Strategy=>BruteForce)
    -- used 0.0146932 seconds
o5 = {}
o5 : List
i6 : randomPoints(J,Strategy=>GenericProjection)
    -- used 0.0146932 seconds
o6 = {{2, -30, -8}}
        \end{verbatim}
    \end{quote}
        }
    \end{example}%
    %
    \vspace{-1em}
    \item[\tt ProjectionAttempts => ZZ]

    When calling the strategy {\tt GenericProjection} or \\
    {\tt HybridProjectionIntersection} from {\tt randomPoints}, this option denotes the number of trials before giving up.  This option is also passed to {\tt randomPoints} by other functions.

@@

    \begin{example}
        We re-compute \cref{BruteForce} this time specifying more attempts.
    {{\small\color{blue}
    ~~
    \begin{quote}
\begin{verbatim}    
i7 : randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 20000)
    -- used 0.84679 seconds
o7 = {{-39, -43, 28}}
    \end{verbatim}%        
\end{quote}\vspace{-1em}%
}}%
    \end{example}%    
%
    %\vspace{1em}
    \item[\tt MaxCoordinatesToTrivialize]

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Even for some simple examples, it could not provide any rational points if the number of trials is not enough.  Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.  A quick comment, while in many examples generic projection works well, there are some examples where it performs quite slowly.  Setting {\tt ProjectionAttempts=>0} in the {\tt Default} strategy is a way to avoid this situation.

    \begin{example}
        We begin with an example over a small field.
    {{\small\color{blue}
    ~~
    \begin{quote}
\begin{verbatim}
i2 : R = ZZ/7[x_1..x_10];
i3 : I = ideal(random(2, R), random(3, R));
o3 : Ideal of R
i4 : time randomPoints(I, Strategy => BruteForce, PointCheckAttempts => 20000)
    -- used 0.0102075 seconds
o4 = {{-2, -1, 3, -2, -2, 2, 0, -2, -2, -2}}
o4 : List
i5 : time randomPoints(I, Strategy => Default)
    -- used 0.085741 seconds
o5 = {{1, -2, 2, 2, -2, 0, 0, -1, -1, 3}}
    \end{verbatim}
\end{quote}\vspace{-1em}
    }}
    \end{example}

\begin{example}
    Now we work over a larger field.
    {{\small\color{blue}
    \begin{quote}
\begin{verbatim}
i6 : S = ZZ/211[x_1..x_10];
i7 : J = ideal(random(2, S), random(3, S));
o7 : Ideal of S
i8 : time randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 2000000)
    -- used 25.3479 seconds
o8 = {{-36, 26, -63, -7, -48, 78, -44, -22, 105, 39}}
i9 : time randomPoints(J, Strategy => Default)
-- used 0.102918 seconds
o9 = {{0, -42, -53, -33, -65, -27, -105, -37, 99, 64}}
o9 : List
i10 : time randomPoints(J, Strategy => LinearIntersection)
-- used 0.0169781 seconds
o10 = {{-51, -11, -34, -33, -65, 19, -86, 50, -44, -97}}
o10 : List
i11 : time randomPoints(J, Strategy => GenericProjection)
-- used 0.364902 seconds
o11 = {{31, -15, 53, 41, 87, 9, -21, -43, -12, 85}}      
    \end{verbatim}
\end{quote}\vspace{-1em}
        }}
    \end{example}

    \begin{example}
        Finally, we allow our functions to extend our field.
        {{\small\color{blue}
        \begin{quote}
        \begin{verbatim}
i12 : time randomPoints(J, Strategy => LinearIntersection, ExtendField => true)
-- used 0.0612306 seconds
            5      4      3      2                                      5  
o12 = {{54, 14a  - 48a  - 56a  + 56a  - 13a + 48, -101, 56, -42, -27, - 26a  -
    -------------------------------------------------------------------------
    4      3      2
    43a  - 70a  - 96a  + 50a - 103, -29, 0, -37}}
        \end{verbatim}
    \end{quote}
        }}
    \end{example}

@@

This function takes a polynomial ring as an input and output the coordinate change map,
i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

@@

\begin{itemize}
\item {\tt Replacement}: Setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
\item {\tt MaxCoordinatesToReplace}: The user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
  This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).		
\end{itemize} 
%\par
%\vspace{-1em}    
\begin{example}  We demonstrate some of these options.
%
  {{\small\color{blue}
  \begin{quote}
\begin{verbatim}
	i3 : R = ZZ/11[x, y, z];
	i4 : randomCoordinateChange(R)
           ZZ
o4 = map(R,--[x, y, z],{4x + 5y - 5z, 3x - 4y - 3z, 4x})                     
           11
                    ZZ
o4 : RingMap R <--- --[x, y, z]
                    11
	i5 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1)
	o5 = | x -x-4y-5z y |
	i6 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Homogeneous => false)
	o6 = | x-3 z-5 -x+3y-4z+2 |
	i7 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Replacement => Binomial)
	o7 = | y x+4z z |
\end{verbatim}
\end{quote}
    }}

@@

\begin{example}	
    We project a curve in 4-space to one 2-space.
    ~~
  {{\small\color{blue}
  \begin{quote}
\begin{verbatim}
i1 : R = ZZ/5[x, y, z, w];
i2 : I = ideal(x, y^2, w^3 + x^2);
i3 : genericProjection(2, I)
            ZZ                                         2          2
o3 = (map(R,--[z, w],{- x - 2y - z, - y - 2z}), ideal(z  - z*w - w ))                   
            5
\end{verbatim}
\end{quote}
    }}
\end{example}\vspace{-1em}
Alternatively, instead of {$I$}, we may pass it a quotient ring.  It will then return the inclusion of the generic projection ring into the given ring, followed by the source of that inclusion.  It is essentially the same functionality as calling {\tt genericProjection(n, ideal R)} although the format of the output is slightly different.

@@

  {{\small\color{blue}
\begin{verbatim}
i1 : R = ZZ/5[x, y, z, w];
i2 : I = ideal(x, y^2, w^3 + x^2);
i3 : genericProjection(2, R/I)	
                            ZZ                          ZZ
                            --[z, w]                    --[z, w]                        
                R           5                           5
o3 = (map(----------------,--------,{2y - w, - y - z}), --------)                       
             2   3    2        4                            4
        (x, y , w  + x )      z                            z
o3 : Sequence
\end{verbatim}}
    }}
\end{comment}

@@

i2 : R = ZZ/5[x, y, z, w];
i3 : I = ideal(x, y^2, w^3 + x^2);
o3 : Ideal of R
i4 :genericProjection(2, I, Replacement => Binomial)
            ZZ                                   4
o4 = (map(R,--[z, w],{x + 2z, - 2y + w}), ideal w )                                     
            5
o4 : Sequence
\end{verbatim}}
    }}
\end{comment}

@@

    o4 : Sequence
	i5 : projectionToHypersurface(R/I)

@@

  {{\small\color{blue}
\begin{verbatim}
i2 : R = ZZ/5[x, y, z];
i3 : I = ideal(random(3, R) - 2, random(2, R))
                3    2       2     3     2              2       2     3
o3 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z )
o3 : Ideal of R
i4 : M = jacobian(I)
o4 = {1} | -x2+2xy-y2+xz+2yz    x-2y+z   |
     {1} | x2-2xy-y2+2xz+2yz+z2 -2x-y-2z |
     {1} | -2x2+2xy+y2+2yz-z2   x-2y-2z  |
             3       2
o4 : Matrix R  <--- R
i5 : extendIdealByNonZeroMinor(2, M, I, Strategy => LinearIntersection)
                3    2       2     3     2              2       2     3
o5 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2   3    2        2    3  
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z , x  + x y + 2x*y  - y  +
-------------------------------------------------------------------
  2       2    3
2x z + y*z  - z )
o5 : Ideal of R
\end{verbatim}}	
    }}	
\end{comment}

@@

\bibliographystyle{alpha}
\bibliography{MainBib}

\vspace{1em}
\end{document}
